---
title: "The Infant Sex-Ratio as a One-sided Measure of Maternal-Infant Health"
author: "Jim Irwin and Jesse McDevitt-Irwin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: ["/home/friend/Research/Bib/bib.bib"]
biblio-style: "apalike"

---

```{r setup, include = FALSE}
library(tidyverse)
```

The sex-ratio of infants (the number of boys over the number of girls, henceforth SR1) is a useful measure of maternal-infant health. The sex-ratio at birth (SRB) of a healthy population is around 106 males per 100 females [@grechUnexplainedDifferencesSex2002; @chahnazarianDeterminantsSexRatio1988], but decreases with maternal stress [@jamesReviewEstablishedSuspected2017]. In infanthood, male infants are more likely to die than female infants [@alkemaNationalRegionalGlobal2014], meaning that the SR1 is negatively related to infant-mortality. The SR1 is available where other measures of population health are not, making it particularly useful for historical settings. 

The sex-ratio of infants is a _one-sided_ measure of maternal-infant health because of discrimination and neglect towards girls. If we observe a low sex-ratio, we can conclude that maternal-infant health conditions are poor. Biologically normal sex-ratios, on the other hand, may come from a healthy population or from sex-discrimination against girls. To demonstrate both the veracity of the SR1 as an indicator, and its one-sidedness, consider the following figure, which plots infant mortality against the SR1 on the country level:



```{r, cars, fig.cap="The SR1 is the ratio of males to females under the age of 1. Infant mortality is in deaths per 1000 live births. Each point is the unweighted average by country across years of SR1 and infant mortality. The data comes from the UN Population Prospects. The black line represents the biologically normal sex-ratio at birth of 105.5.",fig.width=10,fig.height= 5.5, echo=FALSE, message = FALSE, warning=FALSE}
load("~/Research/ssmm/un_sr1.Rdata")
x <-long %>% 
	group_by(Country) %>%  
	summarise(CC=CC,sr1 = mean(sr1,na.rm = T), mort = mean(mort, na.rm=T)) %>%
	ggplot(aes(x=sr1,y=mort))+geom_point(aes(color =CC, label = Country))+geom_smooth(aes(),color="black")+ xlim(.975,1.10)+geom_vline(xintercept=1.055)+
	labs(x= "SR1",y = "Infant Mortality (1000)",title = "Infant Mortality by SR1: 1953-1988")+scale_color_brewer(palette="Set1")
plotly::ggplotly(x, tooltip="label")
```

Below the biologically normal SRB (105.5, represented by a black line), there is a clear, negative relationship between SR1 and infant mortality. For high sex-ratios, we see a weak, positive relationship. Moreover, the bottom-left corner of the figure is empty, which points to the one-sidedness of the measure. If a low SR1, then high infant-mortality. For high SR1's, there is no such relationship. If a high SR1, infant mortality could be high or low.

# Biology and Demography of the SR1

Here we list several key points: 

* Increases in maternal stress, both physical and mental, lead to declines in the sex-ratio at birth.[^1]
* In the absence of sex-discrimination, male infant-mortality is greater than female [@waldronTooYoungGenes1998].
* With greater infant-mortality, the SR1 will drop further from the SRB.
* The effect of increased infant mortality is clearly visible in standard life-tables.
* Increased fertility causes the SR1 to rise, but thsi effect is second-order small relative to the infant-mortality and maternal stress effects.


For a more complete treatment of the biology and demography of the SR1, along with relevant citations, click [here](biology.html).


# Practical Aspects of the SR1 in Historical Data

# Case Studies

## Colonial South Africa

## American Slaves

## 19th Century British Isles

## 19th Century France

The sex ratio of infants in France between 1851-1901 shows evidence of a significant penalty for living in Paris.

```{r,fig.width=10,fig.height= 5.5, echo=FALSE, message = FALSE, warning=FALSE}
load("/home/friend/Research/ssmm/rdata/france_1851_96.Rdata")
france_clean%>% 
  group_by(Year,Set) %>% 
  summarise(sr = ((sum(Male_0, na.rm = T)+sum(Male_1, na.rm = T))/((sum(Fem_0,na.rm = T)+sum(Fem_1,na.rm = T))))) %>% 
  ggplot()+ geom_point(aes(x=Year,y=sr,color=Set),size = 4)
```


# Living Standards and Sex Discrimination: Examples from Spain and India


# References

[^1]: For example, see the seminal work on the Kobe earthquake in Japan [@fukudaDeclineSexRatio1998], and the work of Catalano et.al. (2003, 2005, 2006, and 2008), all of which explore relatively subtle changes in the SRB. For a more extreme example, see that of the Aamjiwnaang nation in Ontario [@mackenzieDecliningSexRatio2005], where toxic chemical exposure has lead to an SRB of 53. 


