---
title: "Infant Sex-Ratios and Maternal-Infant Health"	
subtitle: "A work in progress"
author: "Jesse McDevitt-Irwin[^1] and James R. Irwin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: ["/home/friend/github/sr1/files/0SSMM.bib"]
biblio-style: "apalike"
output: html_document

---

```{r setup, include = FALSE}
library(tidyverse)
```

# Introduction

The infant sex ratio (SR1) offers an important perspective on maternal and infant health (physical and mental) in a population. And observations of sex ratios have great potential as indicators of the well-being of mothers and children (two groups for whom other measures of health and welfare are often lacking).[^2] The sex ratio at birth (SRB) in a healthy population tends towards 106 males per 100 females, but decreases with maternal stress.[^3] And on biological grounds, the infant sex ratio (SR1) falls as mortality rises, because infant mortality tends to be relatively greater among males than females.[^4] Absent sex-based differences in nutrition and care, deleterious living conditions for pregnant women, infants, and nursing mothers in a population will be reflected in low ratios of boys to girls -- at birth, in infancy, and among children.[^5] Elevated child sex ratios are familiar from social science research pursuing Sen's (1990, 1992) call for attention to "missing women" in the context of mortal misogyny; but depressed child sex ratios are simpler indicators of deleterious conditions of life for pregnant women and nursing children. Of particular importance for research on historical populations, evidence on infant and child sex ratios is often available when infant mortality rates and various economic indicators are not.   

However, the infant sex ratio (SR1, henceforth) can also reflect gendered social practices, which discriminate against girls.[^6] When frequent and severe enough, such practices amount to mortal misogyny, which pushes up sex ratios among infants and children.[^7] Given the ubiquity of discimination against girls, when we observe a low infant sex-ratio, we can conclude simply that maternal/infant health conditions were poor. However, apparently healthy child sex ratios could result from mortal misogyny in the context of deleterious living conditions.[^8] Because female infanticide and other forms of mortal misogyny will inflate infant and child sex ratios, the ratios offer a one-sided test of infant and maternal misery. Low infant sex ratios are a strong signal of poor living conditions for infants and mothers, but healthy-looking sex ratios could reflect a combination of poor living conditions and severe forms of discrimination against girls.[^9] When comparing across societies, we must be alert for the potential role of sex-discrimination when interpreting sex-ratio data. 

To make an initial case for the value of the infant sex ratio as a measure of living conditions, consider figure 1, plotting infant mortality against the SR1 on the country level. Below the biologically normal SRB (105.7, represented by a black line), there is a clear, negative relationship between SR1 and infant mortality.[^9] Of particular importance, the bottom-left corner of the figure is empty, which points to the one-sidedness of the measure. If a low SR1, then high infant-mortality. For high SR1's, there is no such relationship. If a high SR1, infant mortality could be high or low. More generally, the plot of infant mortality against the SR1 illustrates the value of low levels of the SR1 as a signal of deleterious living conditions. 

```{r, cars, fig.cap="Sources and notes. Each point is the unweighted average by country across years of SR1 and infant mortality, but the relationship is quite stable across different weightings. The data comes from the UN Population Prospects, available for download [here](https://population.un.org/wpp/Download/Standard/Interpolated/).",fig.width=10,fig.height= 5.5, echo=FALSE, message = FALSE, warning=FALSE, cache=TRUE}
load("~/Research/ssmm/un_sr1.Rdata")
x <-long %>% 
	group_by(Country) %>%  
	summarise(CC=CC,sr1 = mean(sr1,na.rm = T), mort = mean(mort, na.rm=T)) %>%
	ggplot(aes(x=sr1,y=mort))+geom_point(aes(color =CC, label = Country))+geom_smooth(aes(),color="black")+ xlim(.975,1.10)+geom_vline(xintercept=1.055)+
	labs(x= "SR1",y = "Infant Mortality (1000)",title = "Infant Mortality by SR1: 1953-1988")+scale_color_brewer(palette="Set1")
plotly::ggplotly(x, tooltip="label")
```


# Biology and Demography of the SR1

Key points: 

* Increases in maternal stress, both physical and mental, cause declines in the sex-ratio at birth (e.g. [@catalanoNaturalSelectionUtero2013])[^11]  
* In the absence of sex-discrimination, infant-mortality is greater for males than females [e.g. @waldronSexDifferencesInfant1998].
* With greater infant-mortality, the SR1 will drop further from the SRB.
* The effect of increased infant mortality is clearly visible in standard life-tables.
* Increased fertility causes the SR1 to rise, but this effect is second-order small relative to the infant-mortality and maternal stress effects.

For our more complete treatment of the biology and demography of the SR1, along with relevant citations, click [here](biology.html).


# Practical Aspects of the SR1 in Historical Data

* The underlying binomial distribution gives rise to substantial variance in the SR1, so fairly large samples are often required to detect important differences. See [here](Binomial.pdf) for further discussion.
* When census data exhibits male-biased age-heaping at one year old, sex ratios for under age 2 (SR02) or under age 5 (SR05) may serve as effective indicators of well-being. See [here](AgeHeap.html) for an example of male-biased age-heaping in the US census.

# Case Studies

Our work in progress is drawing on a range of census data to explore the usefulness of the SR1 for the study of historical living standards, as follows ...

<a name="southafrica"></a> 

## South Africa

* The SR1 provides striking evidence of the _miserable_ conditions in which black South African lived during the colonial and Apartheid eras. In South Africa in the half-century 1891, the SR1 for Blacks fell in the range 91 to 93 and the SR1 for whites fell in the range 103 to 104. 
* By 1991, the black-white gap in the SR1 had narrowed considerably, with the SR1 for Blacks rising to 101.4, with a much smaller increase in the white SR1, to 104.7. 
* 

In the South African census of 1991, the white SR1 was 104.7, and the black SR1 was 101.4. 

The shrinking of this gap was likely due to the large decrease in infant mortality across South Africa in the post-WW2 period.


<a name="US-slaves"></a>   

## American Slaves

* The SR1 provides clear evidence of the _miserable_ conditions of slaves in the Antebellum South.
* The SR1 of slaves is ~95, while that of whites in the same counties is ~104.
* Since 1860, the SR1 of Black Americans has improved, though it remains below that of whites.

<a name="US-BW20"></a> 

## Black and White in the 20th-Century US
* Sex-ratio evidence points to stagnation of black health in the first half of the 20th century. At census benchmarks before 1950, the black SR2 hovered around 99 points, while the white SR2 crept up from 103 to 104.
* Improvements in the Black SR2, and some narrowing of the racial sex-ratio gap, are evident in the eight decades after 1940. The Black SR2 climbed from 98.9 in 1940 to about 103.1 in 2000.    
* The black-white SR2 gap fell from some 5 percentage points early in the century to just 2 percentage points in 2010.    
* The sex-ratio evidence largely re-enforces current knowledge of racial health inequities in the US.   
* The sex-ratio evidence points to possibilities of errors in current estimates of black infant mortality. Current estimates of the black IMR suggest big improvements in the first half of the 20th century. Sex-ratio evidence suggest instead that the period after 1960 saw the biggest gains in black children's well-being.    

Click [here](US-BW.pdf) for our work on black and white child sex ratios in the 20th-C US. 


<a name="gb-ireland"></a>

## Great Britain and Ireland in the 19th century

* Ireland and Scotland have relatively high SR02's during the mid-19th century, 104.3 and 103.7 respectively. 
* England has a surprisingly low SR02 of 98 in 1841 and 100.6 in 1851.
* Together, these results suggest that a large degree of physical misery came with England's early industrialization and urbanization, relative to its rural neighbors.

<a name="france"></a>

## France in the 19th century

* France has relatively good census data, allowing us to calculate sex-ratios on the _département_ level.
* Paris shows an SR1 consistenly lower than that of either rural _départements_ or smaller cities like Marseilles and Lyon.


For examples of applications of the SR1 to census data from the 19th and 20th century, click [here](cases.html) and see work in progress.

<a name="missing"></a>

# The SR1 and Missing Girls

Given the historical availability of the population sex-ratios, a natural question would be whether the SR1 might also shed light on the question of missing women. As stated above, a low SR1 is unambiguous and implies poor maternal-infant conditions. A high SR1, e.g. 105, on the other hand, might be caused by good maternal-infant conditions, or by female infanticide and neglect. Looking only at the SR1, these two effects cannot be disentangled. However, by comparing across age groups, we may be able to detect excess female mortality. See [here](sex_discr.html) for a treatment of the SR1 and missing girls.


<a name="references"></a>

# References

[^1]: Authors' information: Corresponding author is Jesse McDevitt-Irwin, PhD Student in Sustainable Development at Columbia University (jrm2275@columbia.edu). James R. Irwin is retired Professor of Economics, Central Michigan University.

[^2]: As discussed below, high values of child sex ratios are familiar from social science research on "missing women". Our attention to *low* sex ratios among infants and children is an extension of recent research on the sex ratio at birth (SRB) as a population-level health indicator: e.g. [@jamesCanSexRatios2018], [@grechSexRatioBirth2016]. Earlier work identified reductions in the SBR as a result of stressful events (floods, earthquakes, severe family health events, and the 9-11 attacks: see Lyster (1965), [@fukudaDeclineSexRatio1998], [@hansenSeverePericonceptionalLife1999], [@catalanoSexRatiosCalifornia2005]). 

[^3]: For a healthy SRB of 106 see e.g. [@maconochieSexRatiosAre1997], reporting (p. 353) a SRB of 1.058 in over 1/2 million births in Scotland for the period 1975-1988. For the SRB decreasing with maternal stress, see e.g. [@westSystematicSearchFactors2020]. See also [@jamesReviewEstablishedSuspected2017], on debate over precise mechanisms and evolutionary implications, because of which the SRB "remains one of the most longstanding unresolved problems in the biological sciences" (2017:50); further to this point see [@brucknerSelectionUteroPopulation2018]. Much of the SBR the research explores small/subtle changes in the SRB (less than 2 percentage-points), but [@catalanoExogenousShocksHuman2006] find a 5-point drop in the SBR in New York due to the stress of 9-11 attack. In a more extreme case, in the Aamjiwnaang nation in Ontario  [@mackenzieDecliningSexRatio2005], toxic chemical exposure led to an SRB of just 53.  

[^4]: See e.g. [@waldronSexDifferencesHuman1983], pp. 324-25. See also [@hammoudStudiesFetalInfant1965]. An early and convenient statement of male-biased infant and child mortality is offered in the 1861 Census of Scotland: "At birth 106 males appear for every 100 females ...; yet, by the law of nature, -- a law seen in still more powerful operation even in intra uterine life -- the tendency to succumb under disease is so much greater in the male than in the female, that by about the 18th year of life the proportion of the sexes becomes equalized, 100 males being alive at that age for every 100 females".

[^5]: Although female-biased mortality is a possibility for a few diseases ([@garenneSexistDiseases1998], p. 176), on the whole it seems clear that it requires sex-based discrimination against girls in nutrition and care to generate a rising sex ratio in a population (where the population is large enough that random fluctuations in the sex ratio cancel out).

[^6]: We cannot find any populations with gender-based discrimination against males, so we note but ignore the possibility that low infant sex ratios could reflect discrimination against males in an otherwise healthy population. 

[^7]: Post-ultra-sound, misogyny may be reflected in an elevated sex ratio at birth.

[^8]: For example, consider the Punjab province of British-ruled India in 1891. The infant sex ratio there was 103, a healthy-looking value for the time. However, the child sex ratios were 108.0 for children aged 1-5 and 118.4 for those aged 5-9. Our understanding is that female infanticide, and not favorable living conditions, is the source of the healthy-looking SR1 (see [@guptaSelectiveDiscriminationFemale1987], pp. 88,93, on infanticide in the Punjab). More generally, the pattern of dramatically increasing sex-ratios with age is evidence of misogynistic gender-discrimination. Further to this point, see [below](#missing), and our discussion of [Sex-ratios and Missing Girls](sex_discr.html)

[^9]: And extreme forms of gender discrimination can result in infant and child sex ratios inflated well in excess of the biological norm of 106. Simply, any large population with a child sex ratio of 108 or higher is suspected of mortal misogyny -- female infanticide or life-ending deprivation of girls.

[^10]: Note that at low levels of infant mortality, the female survival advantage will have a negligible impact, and the SRB and SR1 will be about the same.

[^11]: [@helleEvolutionaryEcologyHuman2009] provides a useful listing of relevant research finding that "stressful conditions caused by both abiotic and biotic hazards and economic deprivation ... skew human birth sex ratio towards women", citing Lyster 1974; Fukuda et al. 1998; Hansen, Møller & Olsen 1999; Catalano 2003; Stein, Zybert & Lumey 2004; Catalano & Bruckner 2005; Catalano et al. 2005a, b, c, 2006; Kemkes 2006; Obel et al. 2007; Saadat 2008. See also above, note 3. And see [@winstonInfluenceSocialFactors1931] for an earlier statement drawing a causal link from affluence to a high SBR. 
