---
title: "The Infant Sex-Ratio as a One-sided Measure of Maternal-Infant Health"
author: "Jim Irwin and Jesse McDevitt-Irwin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include = FALSE}
library(tidyverse)
```

The sex-ratio of infants (the number of boys over the number of girls, henceforth SR1) is a useful measure of maternal-infant health. The sex-ratio at birth (SRB) in a healthy population tends towards 106 males per 100 females, but decreases with maternal stress. In infanthood, male infants are more likely to die than female infants, meaning that increasing infant-mortality also drives down the SR1. The sex-ratio of infants is available where other measures of population health are not, making it particularly useful for historical settings. 

The sex-ratio of infants is a _one-sided_ measure of maternal-infant health because of discrimination and neglect towards girls. If we observe a low sex-ratio, we can conclude that maternal-infant health conditions are poor. But if we observe a biologically normal sex-ratio, we cannot determine (based on that information alone) whether the population is healthy, or unhealthy but with sex-discrimination driving up the sex-ratio. To demonstrate both the veracity of the SR1 as an indicator, and its one-sidedness, consider the following graph, which plots infant mortality against the SR1 on the country level:



```{r, cars, fig.cap="A scatterplot.",fig.width=10,fig.height= 5.5, echo=FALSE, message = FALSE, warning=FALSE}
load("~/Research/ssmm/un_sr1.Rdata")
x <-long %>% 
	group_by(Country) %>%  
	summarise(CC=CC,sr1 = mean(sr1,na.rm = T), mort = mean(mort, na.rm=T)) %>%
	ggplot(aes(x=sr1,y=mort))+geom_point(aes(color =CC, label = Country))+geom_smooth(aes(),color="black")+ xlim(.975,1.10)+geom_vline(xintercept=1.055)+
	labs(x= "SR1",y = "Infant Mortality (1000)",title = "Infant Mortality by SR1: 1953-1988")+scale_color_brewer(palette="Set1")
plotly::ggplotly(x, tooltip="label")
```

For low levels of the SR1, there is a clear, negative relationship with infant mortality. For high sex-ratios, we see a small, positive relationship. Moreover, the bottom-left corner of the figure is empty, which points to the one-sidedness of the measure. If a low SR1, then high infant-mortality. For high SR1's, there is no such relationship. If a high SR1, infant mortality could be high or low. 

# Biology and Demography of the SR1

Key points:

* Increases in maternal stress, both physical and mental, lead to declines in the sex-ratio at birth.
* This effect happens through spontaneous miscarriage, which disproportionately affects male fetuses.
* In the absence of sex-discrimination, the SR1 will be less than the SRB, because male infants are more likely to die.
* With greater infant-mortality, the SR1 will drop further from SRB.
* The effect of increased infant mortality is clearly visible in standard life-tables.
* Increased fertility causes the SR1 to rise, as the median age of infants in a population grows larger. 
* This fertility effect, by life-table calculations, is second-order small relative to the infant-mortality and maternal stress effects.


For a more complete treatment of the biology and demography of the SR1, click [here](biology.html).